### 2025-07-04 causal-inference.qmd
20250704 - edits for grammar and clarity

So, our goal in working toward causal inference as scientists is to create reasonable approximations to this impossible counterfactual [@Shadish2002].
For instance, if using a [between-subject design](#sec-betweenSubject), we want the two groups to be equivalent in every possible way except whether or not they receive the treatment, so we might stratify each group to be equivalent in terms of age, weight, position, experience, skill, etc.
Or, we might test the same people using an A-B-A-B [within-subject design](#sec-withinSubject).
In an A-B-A-B [within-subject design](#sec-withinSubject), players receive no treatment at baseline (timepoint 1: game 1), receive the treatment at timepoint 2 (game 2), receive no treatment at timepoint 3 (game 3), and receive the treatment at timepoint 4 (game 4).
Neither of these approximations is a true counterfactual.
In the [between-subject design](#sec-betweenSubject), the players differ between the two groups, so we cannot know how the individuals who received the treatment would have performed if they had actually not received the treatment.
In the A-B-A-B [within-subject design](#sec-withinSubject), the players are the same, but they timepoints that they receive or do not receive the treatment differ, and there can be [carryover effects](#sec-withinSubject) from one condition to the next.
For instance, consuming sports drinks before game 2 might also help them be better hydrated in general, including, for subsequent games.
Thus, we cannot know how a player would have performed in game 1 with treatment or in game 2 without treatment, etc.
Nevertheless, it is important to be aware of the counterfactual and to engage in counterfactual reasoning to consider what would have happened if the supposed cause had not occurred.
Considering the counterfactual is important for designing closer approximations to the counterfactual in studies for stronger research designs and stronger causal inference.

>>>>

So, our goal in working toward causal inference as scientists is to create reasonable approximations to this impossible counterfactual [@Shadish2002].
For instance, if using a [between-subject design](#sec-betweenSubject), we want the two groups to be equivalent in every possible way except whether or not they receive the treatment, so we might stratify each group to be equivalent in terms of age, weight, position, experience, skill, etc.
Or, we might test the same people using an A-B-A-B [within-subject design](#sec-withinSubject).
In an A-B-A-B [within-subject design](#sec-withinSubject), players receive no treatment at baseline (timepoint 1: game 1), receive the treatment at timepoint 2 (game 2), receive no treatment at timepoint 3 (game 3), and receive the treatment at timepoint 4 (game 4).
Neither of these approximations is a true counterfactual.
In the [between-subject design](#sec-betweenSubject), the players differ between the two groups, so we cannot know how the individuals who received the treatment would have performed if they had actually not received the treatment.
In the A-B-A-B [within-subject design](#sec-withinSubject), all players receive the same sequence of alternating conditions—no treatment, treatment, no treatment, treatment—across four timepoints.
Although the A-B-A-B [within-subject design](#sec-withinSubject) keeps participants constant, the specific occasion of measurement may differ from participant to participant (thus introducing time or context of measurement as a potential [confound](#sec-causalDiagramConfounding)), and there can be [carryover effects](#sec-withinSubject) from one condition to the next.
For instance, consuming sports drinks before game 2 might also help the player be better hydrated in general, including for subsequent games.
Thus, we cannot know how a player would have performed in game 1 with treatment or in game 2 without treatment, and so on.
Nevertheless, it is important to be aware of the counterfactual and to engage in counterfactual reasoning to consider what would have happened if the supposed cause had not occurred.
Considering the counterfactual is important for designing closer approximations to the counterfactual in studies for stronger research designs and stronger causal inference.
